Sampling Strategies in Decisions from Experience

### 2022-05-19 manuscript.Rmd
update section 1: sequential sampling framework

\noindent
\fbox{
    \parbox{\textwidth}{
        \textbf{\hypertarget{box}{Box 1.}} 
        \textit{Reliance on Small Samples in DFE.}
        The binomial distribution of an outcome is the probability distribution of the number of times an outcome occurs in a sample.
        According to the \textit{de Moivre-Laplace theorem} (see, e.g., \protect\hyperlink{ref-georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015}{Georgii, 2015}), the binomial distribution can be approximated by a symmetric normal distribution given a large sample.
        That is, the relative frequency with which an outcome occurs in a large sample is predicted to correspond to the objective probability of the outcome most of the time and is not predicted to be lower more often than higher or vice versa.
        However, for small samples, the binomial distribution associated with a small-probability outcome is positive skewed, causing the relative frequency with which this outcome occurs in the sample to be smaller rather than larger than its objective probability.
        That is, small-probability outcomes tend to be underrepresented in small samples.
        An explanation for the as-if underweighting of rare outcomes pattern may thus build on the assumption that the mind is sensitive to the information provided by small samples (\protect\hyperlink{ref-hertwigDecisionsExperienceEffect2004}{Hertwig et al., 2004}; \protect\hyperlink{ref-plonskyRelianceSmallSamples2015}{Plonsky et al., 2015}; see also \protect\hyperlink{ref-erevChoicePredictionCompetition2010}{Erev et al., 2010}, \protect\hyperlink{ref-erevAnomaliesForecastsDescriptive2017}{2017}, for the performance of models assuming reliance on small samples in prediction competitions).
    }
}
\newline

@@

\begin{figure}[t]
\caption{Possible Sampling and Decision Strategies in the Sampling Paradigm and Their Impact on the Final Choice}
\includegraphics{manuscript_files/sampling-paradigm} \label{fig:sampling-paradigm}
\raggedright \textit{Note.} Add note
\end{figure}

@@

4. *No discussion of choice problems:*
Assuming representative sampled frequencies, summary and round-wise decisions only predict different choices, if the prospect with the larger EV is not also more likely to produce a higher outcome most of the time ((see Fig.\ \@ref(fig:sampling-paradigm)).
In consequence, the structure of the choice problems---i.e, whether a small-probability outcome exists and how desirable it is---can quite dramatically shape how the (interplay of) strategies differ in the choices and as-if weighting patterns they produce.
Focusing only on problems of such a structure may provide a parsimonious but narrow-in-scope interpretation.

The current paper aims to extend the work of @hillsInformationSearchDecisions2010 and to address the outlined problems.
The rest of the paper is structured as follows:
In the first section, the current approach is described

This paper presents a couple of models that integrate two proposed sampling and decision strategies [see @hillsInformationSearchDecisions2010] into the computational framework of sequential sampling models of decision-making [see, e.g., @markantModelingChoiceSearch2015].
More specifically, controlling for sampling error---i.e., considering the sampled relative frequencies rather than the latent objective probabilities---, this paper seeks to investigate in a computational analysis whether and how the potential interplay of sampling and decision strategies shapes DFE and comes to imply choice patterns and functional forms of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992] that are consistent with an as-if underweighting of rare outcomes pattern.^[<!--Open footnote-->
CPT's value function is also considered in this paper.
However, the focus is on the weighting function, as it is commonly used to describe an as-if overweighting and underweighting of rare outcomes, and the description-experience gap [see, e.g., @kellenHowVariantAre2016; @regenwetterConstructbehaviorGapBehavioral2017; but see @hertwigConstructbehaviorGapDescription2018].
<!--Close footnote-->]

Thereafter, a simulation study is presented, where the proposed models are used to simulate the sampling and accumulation processes for a set of choices between a safe and a two-outcome prospect.
Considering choice rates and parameter estimates of CPT, the computational analysis shows that ... which translate to characteristic patterns in the psychoeconomic functions of CPT.
[...]
The paper closes with a discussion and conclusion.
Note that all materials, including the can be retrieved from the accompanying `GitHub repository`.

# Explaining Decision Making in the Sampling Paradigm

<!-- Formal Definition of Prospects / Notation -->

<!--Since there is an uncountable number of choices people confront in their daily life, behavioral decision research routinely abstracts from choices between particular alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two prospects, it studies a case which is counterfactual in that it omits the many particularities of each choice situation, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess.-->
With the choice between at least two prospects, behavioral decision research studies a case which is counterfactual in that it omits the many particularities of each choice situation, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess.
These fundamental properties are considered the $n$ possible outcomes of a choice alternative, denoted by the set $\{x_i\}_{i \in \{1, ..., n\}}$, and the probabilities with which these outcomes occur following the choice of the alternative, denoted by the set $\{p(x_i)\}$.
That is, each choice alternative that is fully described by a set of outcome-probability pairs $\{(x_i, p(x_i))\}$ satisfying $\sum_{i = 1}^n p(x_i) = 1$ is a prospect.
Note that a prospect can be conceptualized as the distribution $P$ of a discrete random variable $X$, where each outcome in $\{x_i\}$ denotes a possible value that $X$ can take.
In other words, then, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
\{P^1, ... , P^k\}
\; .
\tag{1}
$$
<!-- Modeling steps and assumptions -->

Since it is only through their fundamental properties, or combinations thereof, that any distinction between the prospects can be made, systematic---as opposed to random---choices between prospects must be based on the assessment of differences in their properties.
However, only rarely in peoples' daily life, the inputs to the decision-making process take the form of explicit descriptions of all outcome-probability pairs.
In such cases, people would make DFD, since the information from which the mind learns about the properties of prospects are complete descriptions thereof.
Yet, rather often, people make DFE, where the mind can learn about the latent properties only by experiential sampling over time [@hertwigDescriptionexperienceGapRisky2009, p. 517].
Accordingly, for the choice between $1, ..., k$ prospects, the inputs to the decision-making process are generated by a set of prospect-specific stochastic processes
$$
\{\{X^1_t\}, ... , \{X^k_t\}\}
\; ,
\tag{6}
$$
where each $\{X_{t}\}_{t \in \mathbb N}$ is a collection of independent and identically distributed (i.i.d.) random variables.^[Note that it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. However, in the sampling paradigm, i.i.d. random variables are commonly assumed.]
The inputs to DFE are then the realizations of the random variables in (6).

@@

# Current Approach: Sampling Strategies and the Accumulation of Different Types of Evidence

<!-- How to overcome the problems + Evidence accumulation -->

The mind may use these realizations---i.e., the sampled outcomes---to first infer the latent properties and then proceed by selecting a property, or combinations thereof, to assess differences among the prospects and make a respective choice.
However, the mind may also approximate such a process by directly accumulating evidence about the differences in latent properties from the sampled outcomes.
To this end, sequential sampling models of decision making [e.g., @busemeyerDecisionFieldTheory1993] assume that the outcomes of prospects are sequentially sampled and integrated into dynamic decision variables and that a choice is made once a decision variable exceeds a threshold in favor for one of the prospects.
This class of process models thereby provides an algorithmic link between the actual sampling process and the choices in DFE, and has shown to explain some of the robust deviations from EV maximization [see @bhatiaSequentialSamplingParadoxes2014; @busemeyerDecisionFieldTheory1993], including the as-if underweighting of rare outcomes in the sampling paradigm [see @markantModelingChoiceSearch2015].
Moreover, these models implement speed-accuracy trade-offs, with the amount of evidence that must be sampled in favor for a prospect increasing with thresholds, leading to more (less) accurate but slower (faster) choices for high (low) thresholds.

@@

## Models

@@

### Round-Wise Integration Model

@@

### Summary Integration Model

@@

# Model Predictions

>>>>

\noindent
\fbox{
    \parbox{\textwidth}{
        \textbf{\hypertarget{box}{Box 1.}} 
        \textit{Reliance on Small Samples in DFE.}
        The binomial distribution of an outcome is the probability distribution of the number of times an outcome occurs in a sample.
        According to the \textit{de Moivre-Laplace theorem} (see, e.g., \protect\hyperlink{ref-georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015}{Georgii, 2015}), the binomial distribution can be approximated by a symmetric normal distribution given a large sample.
        That is, the relative frequency with which an outcome occurs in a large sample is predicted to correspond to the objective probability of the outcome most of the time and is not predicted to be lower more often than higher or vice versa.
        However, for small samples, the binomial distribution associated with a small-probability outcome is positive skewed, causing the relative frequency with which this outcome occurs in the sample to be smaller rather than larger than its objective probability.
        That is, small-probability outcomes tend to be underrepresented in small samples.
        An explanation for the as-if underweighting of rare outcomes pattern may thus build on the assumption that the mind is sensitive to the information provided by small samples (\protect\hyperlink{ref-hertwigDecisionsExperienceEffect2004}{Hertwig et al., 2004}; \protect\hyperlink{ref-plonskyRelianceSmallSamples2015}{Plonsky et al., 2015}; see also \protect\hyperlink{ref-erevChoicePredictionCompetition2010}{Erev et al., 2010}, \protect\hyperlink{ref-erevAnomaliesForecastsDescriptive2017}{2017}, for the performance of models assuming reliance on small samples in prediction competitions).
    }
}
\newline
(ref:references-box-1) @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015; @hertwigDecisionsExperienceEffect2004; @erevChoicePredictionCompetition2010; @plonskyRelianceSmallSamples2015; @erevAnomaliesForecastsDescriptive2017

@@

\begin{figure}[t!]
\caption{Possible Sampling and Decision Strategies in the Sampling Paradigm and Their Impact on the Final Choice}
\includegraphics{manuscript_files/sampling-paradigm} \label{fig:sampling-paradigm}
\raggedright \textit{Note.} In the sampling paradigm, people are offered the choice between at least two prospects (here: violet vs. orange), whose exact properties (outcomes and probabilities) remain hidden. To learn about the properties, people can, as much as they want, sample the prospects' outcomes according to their probabilities. Sampling strategies: Refer to the sequential pattern with which people sample the outcomes from both prospects over time. People may switch infrequently (comprehensive sampling) or frequently (piecewise sampling) or something in between. Decision strategies: Refer to the way people integrate the sampled outcomes in order to evaluate the prospects. People may integrate all outcomes stemming from the same prospect in order to compute and compare summary scores (summary integration/decision); or they integrate and compare outcomes of the two prospects over repeated rounds (round-wise integration/decision).  
\end{figure}

@@

4. *No discussion of choice problems:*
Assuming representative sampled frequencies, summary and round-wise decisions only predict different choices, if the prospect with the larger EV is not also more likely to produce a higher outcome most of the time (see Fig.\ \@ref(fig:sampling-paradigm)).
In consequence, the structure of the choice problems---i.e, whether a small-probability outcome exists and how desirable it is---can quite dramatically shape how the (interplay of) strategies differ in the choices and as-if weighting patterns they produce.
Focusing only on problems of such a structure may provide a parsimonious but narrow-in-scope interpretation.

The current paper aims to extend the work of @hillsInformationSearchDecisions2010 and to address the outlined problems.
The rest of the paper is structured as follows:
In the first section, the current approach and its theoretical commitments are briefly outlined.
The paper starts from the proposition that systematic choices in DFE must be based on the assessment of differences in the latent properties of prospects (e.g., differences in EV or differences in the probability to produce a higher outcome most of the time).
Sequential sampling models of decision making are considered as a theoretical framework that can be used to explain how the mind samples and accumulates evidence about such differences over time, and---importantly---how the mind stops sampling once an evidence threshold is reached.
In the section that follows, two models are specified, each of which integrates the continuum of sampling strategies and one of the proposed decision strategies into the computational framework of sequential sampling models.
The expected behavior of the models are outlined.
Thereafter, a simulation study is presented, where the proposed models are used to simulate the sampling and accumulation processes for a set of choices between a safe and a two-outcome prospect, the latter of which may or may not possess a small-probability outcome of different rank/desirability.
Controlling for sampling error---i.e., considering the sampled relative frequencies rather than the latent objective probabilities---, this paper then seeks to investigate in a computational analysis whether and how the potential interplay of sampling and decision strategies shapes DFE and comes to imply choice patterns and functional forms of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992] that are consistent with an as-if underweighting of rare outcomes pattern.^[<!--Open footnote-->
CPT's value function is also considered in this paper.
However, the focus is on the weighting function, as it is commonly used to describe an as-if overweighting and underweighting of rare outcomes, and the description-experience gap [see, e.g., @kellenHowVariantAre2016; @regenwetterConstructbehaviorGapBehavioral2017; but see @hertwigConstructbehaviorGapDescription2018].
<!--Close footnote-->]
It is shown that within a sequential sampling framework, the interplay of sampling and integration/decision strategies produces systematically different choice patterns that depend on the structure of the choice problems and translate to distinct signatures in CPT's weighting and value function.
The simulated choices and estimated CPT parameters can be understood as model-implied data and data models, which remain to be tested against empirical data.
The paper closes with a brief discussion and conclusion.
Note that all materials---i.e., code and data---underlying this paper, including instructions on how to reproduce this manuscript, can be retrieved from the accompanying `GitHub` repository.

# Sampling and Decision Strategies Within a Sequential Sampling Framework

Because there is an uncountable number of decisions people confront in their daily life, behavioral decision research routinely abstracts from choices between particular alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two prospects, it studies a case which is counterfactual in that it omits the many particularities of each choice situation, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess.
These fundamental properties are considered the $n$ possible outcomes of a choice alternative, denoted as $x_1, ..., x_n$, and the probabilities with which these outcomes occur following the choice of the alternative, denoted as $p_X(x_1), ..., p_X(x_n)$. 
That is, each choice alternative that is fully described by a set of outcome-probability pairs, $\{(x_i, p_X(x_i))\}_{i \in \{1, ..., n\}}$, satisfying $\sum_{i = 1}^n p_X(x_i) = 1$, is a prospect.
It is only through these fundamental properties, or combinations thereof, that any necessary distinction between the prospects can be made.
Hence, systematic---as opposed to random---choices between prospects must be based on the assessment of differences in their properties.

Rarely in peoples' daily life, the information provided about the prospects take the form of explicit and complete descriptions of all their outcome-probability pairs.
In such rare cases, people would make DFD and can---strongly simplified---proceed by selecting a property or combinations thereof to assess differences among the prospects and make a respective choice.
In stark contrast and rather often, people are required to make DFE, where the mind can learn about the latent properties only by "experiential sampling [of outcomes] over time" [@hertwigDescriptionexperienceGapRisky2009, p. 517, square bracket added].
The sampling paradigm is just one paradigm that can be used to study DFE and allows people to sample the prospects' outcomes without costs and without constraints regarding the sequential pattern and the amount of sampled outcomes.

How do people arrive at a final choice in the sampling paradigm? 
Sequential sampling models of decision making under risk (DFD) and/or uncertainty [DFE, see @markantModelingChoiceSearch2015, who explicitly consider DFE and the sampling paradigm; see also, e.g., @bhatiaSequentialSamplingParadoxes2014; @busemeyerDecisionFieldTheory1993] assume that the decision-making process takes the form of an accumulation of evidence for the prospects over time.
The process is assumed to stop---i.e., a choice is made---once the accumulated evidence for one of the prospects reaches a certain threshold.
More specifically, for DFE, it is assumed that the outcomes of prospects are sequentially sampled and integrated into dynamic decision variables and that a choice is made once the value of a decision variable exceeds a threshold.
How exactly this integration is carried out and what the value of a dynamic decision variable represents, is a theoretical question---and thus a matter of the exact specification of the model---largely concerning the way .

This class of process models thereby provides an algorithmic link between the actual sampling process and the choices in DFE, and has shown to explain some of the robust deviations from EV maximization [see @bhatiaSequentialSamplingParadoxes2014; @busemeyerDecisionFieldTheory1993], including the as-if underweighting of rare outcomes in the sampling paradigm [see @markantModelingChoiceSearch2015].
Moreover, these models implement speed-accuracy trade-offs, with the amount of evidence that must be sampled in favor for a prospect increasing with thresholds, leading to more (less) accurate but slower (faster) choices for high (low) thresholds.

@@

@@

# Models

<!--Note that a prospect can be conceptualized as the distribution $P$ of a discrete random variable $X$, where each outcome in $\{x_i\}$ denotes a possible value that $X$ can take.
In other words, then, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
\{P^1, ... , P^k\}
\; .
\tag{1}
$$
-->

@@

## Round-Wise Integration Model

@@

## Summary Integration Model

@@

# What Model Behaviors Are Expected?
